# Replicability

The key to a good, robust analysis is replicability. When we say 'replicability', we mean two things:

* That your results can be replicated in other populations and in other settings (i.e. typical 'scientific' replicability)
* That you can easily share your code and method with others who can verify your outputs

If your project is replicable, then it's likely to have fewer issues, make fewer dodgy assumptions, and rely less on the idiosyncracies of your environment or coding practice. That doesn't mean that everything that you do that can be replicated is immediately correct, but it's a useful credential to have.

For now, we're going to focus on how you can make your *code* more replicable. By that I mean, how you can share your results and your code with others either in your business or instution or even in the open source community in general.

## Good practice

There are a few things you can do to make your work immediately more readable for others:

### Use a consistent naming convention

Take your pick. Everyone has their preference and that's okay. If you like camelCase then go with camelCase. If you like the `_` approach, then go for that. The most important part of your naming convention however is that it's stable. Don't name some of your functions `like_this` and the rest `likeThis`. Not only does it make it harder to read, but every time you do it, a kitten dies. So be consistent.

### Name your variables as nouns and your functions as verbs

Functions do and variables and objects are. Almost all languages share this distinction with verbs and nouns, so utilise that natural divide to improve how you name your functions and variables. Aim to give your functions names that suggest that they `do` something; and bonus points if you can give it a name that's a verb and also gives a decent description of what the function does. When you name your variables, give them meaningful noun-like names. Say you're doing some climate analysis, a good name for the variable that holds the average rainfall in a year might be `avg_yearly_rainfall` whilst a function that converts a temperature in Celsius to Fahrenheit might be `convert_degrees()`.

### Use functions where you can

R is a functional language and so using and creating functions is at the very heart of programming in R. Rather than relying on R scripts that need to be run in their entirety to produce your output, by creating functions and using them in your analysis, you can more easily scale your project. For example, imagine your scope is initially to produce a report for the year. Then, after your done, your manager is so impressed with your report, they want you to do the same thing for the last 10 years. If you've used a couple of R scripts and you haven't written any functions, then this is likely going to involve copying and pasting a whole lot of code and changing the years. Instead, if you use functions to perform your analysis, then creating reports for multiple years can be as simple as changing your dataset.

As a rough rule of thumb, if you find yourself copying and pasting your R code more than twice to do very similar things, you should probably be using a function.

### State your dependencies

One of the great things about R is the number of packages that are available that let you do all sorts of weird and wonderful things. Unfortunately, because there are so many great packages, it's unlikely that the person you're sharing your code with will have them all installed. If you're sending someone a script, then best practice is to include all the packages you use at the top of your script like this:

```{r, eval = FALSE}
library(ggplot2)
library(dplyr)
```

An extra step which few people do but can be very helpful is to still prepend your functions with which package they came from like this `dplyr::mutate()`. Not only does this avoid any namespace conflicts where two packages might have functions with the same name and you don't end up using the one you think you are because of the order of your `library()` calls, but it also makes it infinitely easier for anyone reading your code to find the package that a function comes from.
Admittedly, this is overkill in a sense because we've already told R which packages we're using with our `library()` calls and R has loaded in the objects from that package, but this practice can really improve the readability of your code.

Later we'll look at designing our project as a package and packages have a different way of stating dependencies for the user, so this is primarily for the case where you're just sending a script or two to someone.

## Documentation

When you're working on a project, never think that you'll remember everything when you come back to it. You won't. Instead, imagine you're always writing for code for someone who's never seen it before, because, trust me, when you come back to a project after 6 months you'll have absolutely no idea what you're looking at.

At the heart of writing code that's easy to understand is documentation. Here we'll talk about the two main types of documentation.


### Function documentation

To get a package on CRAN, all the functions that your package exports needs to be documented. So if you type, say, `?dplyr::mutate()` into the console and hit Enter, you'll be taken to the Help page for the `mutate` function. This ensures that the package users are not left guessing what a parameter is supposed to be, or what they're going to get returned from the function.

Even if you're not ever planning on submitting your work to CRAN, function documentation is extremely useful. The `roxygen` package makes this documentation as easy as possible for package developers, and you can use the same principles in your analysis.

At the very least, you should be documentating the input variables (the `@param` tag in `roxygen`), your return value(s) (the `@return` tag) and maybe an example or two (the `@examples` tag). This will make explaining your code to someone else or relearning it yourself infinitely easier.

### Long-form documentation

Whilst understanding what each of your functions does is important, it's also important to document how all of the little pieces fit together. To acheive this, it's a good idea to write some long-form documentation, like a README or a vignette.

Long form-documentation is often written in something called RMarkdown. RMarkdown is a spin on [markdown](https://en.wikipedia.org/wiki/Markdown) that allows you to embed and run R code when generating a document. This can be really useful for explaining your process and sharing your workings.

#### READMEs

READMEs act as an entrypoint for anyone that stumbles across your package. It should be in the root of your project directory, and should be written in markdown or plain text. Anyone should be able to read your README and understand what the project is doing. I won't go into exactly what should be included because it will depend on the type of project your doing, but you know a good README when you see it.

If you also use a repository system like GitHub, your README will act as the homepage and so should always be present.

#### Vignettes


Vignettes are less defined that READMEs. Vignettes should be an in-depth form of documentation for a concept in your project. For example, say that your project is looking at the level of data quality of various open source APIs. You might have a few different vignettes covering different important concepts in your project:

1. Background
    + This would cover why the how the project came to be, and what the ultimate goal is.
2. Selection of APIs
    + Why were certain APIs chosen and others excluded? That would be explained here.
3. Methodology
    + How are you assessing data quality? Here you would explain your different tests and standards.

Whilst a README serves as a general introduction to the project, vignettes more often serve as in-depth descriptions of certain aspects of the project. Like READMEs, vignettes should be written in markdown (or more likely RMarkdown).


## Testing

One thing I rarely see in analysis projects are unit tests. Unit tests are test cases that help ensure that your functions are performing as you expect. For example, let's say you have a function like this:

```{r}
trim_hyphen <- function(txt) {
  gsub("-", "", txt)
}
```

This seems pretty straightforward - it strips hyphens out of text strings. But what happens if you pass this function a `NULL` value? Or a list? Writing test cases to cover these events can help you limit the number of unexpected behaviours and bugs that can often ruin an analysis. I've fallen victim to thinking my analysis was valid before discovering that there was an aggregation or modelling error due to edge cases more than I care to admit.

### `{testthat}`

`{testthat}` is a package designed primarily for testing packages, and when we look in the next chapter at laying out your project like a package, that will make a lot more sense. But your analysis doesn't have to be a package. You just need to write test cases, put them in a script and a folder and run `testthat::test_local("path/to/your/test/director")`.

We won't go into the intricacies of the `{testthat}` package here, but essentially `{testthat}` relies on expectations. That is, the is an expected behaviour, and the test should pass if that expectation is met. If not, the test should fail. Let's test our `trim_hyphen` function:

```{r}
library(testthat)

test_that("trimming hyphens errors on NULL",
                    {
                      expect_error(trim_hyphen(NULL))
                      
                    })
```

Here we setup our test context - that we're checking that when we trim hyphens, we get an error if we provide a NULL value. Then we provide our expectation - we expect an error when we call the function with a NULL value. But our expectation isn't met - there isn't an error. So we need to either change the test case, or the function. In this example, let's say that we do want the function to fail when there's a NULL value, so we'd change the function:

```{r}
trim_hyphen <- function(txt) {
  if (!is.null(tst)) {
    gsub("-", "", txt)
  } else {
    stop ("txt cannot be NULL")
  }
}
```

Now when we run the test case:

```{r}
test_that("trimming hyphens errors on NULL",
                    {
                      expect_error(trim_hyphen(NULL))
                      
                    })
```

The test passes!

### Defensive programming

Defensive programming is a type of design whereby you prioritise safe or continued use of a piece of software in variable and unforeseen circumstances. Many people who practice this type of programming will first define the test cases for their function - how it should operate in any given circumstance. Only after considering all the possibilities would the developer actually then write the definition of the function.

Personally, I don't often advocate this approach to programming in interactive data analysis projects, but it can sometimes be a very interesting and useful approach. By first defining how the function or function(s) should behave in many different circumstances, you can better understand how your analysis or system is going to function outside of the 'happy path'.

So while I wouldn't adapt this practice for all your projects, keep the principles of defensive programming in mind when you're writing a particularly complex or temperamental function.